---
categories:

date: "2020-09-20"
description: American Society's Social Behaviour in 2016
draft: false
keywords: ""
slug: social_american_society
title: How do Americans behave on Social Media?

---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(infer)
library(ggrepel)
library(knitr)
library(ggpubr)
library(scales)
library(tidytext)
```


# General Social Survey (GSS)

**How does the American Society relate to social media, technology and education?**

The [General Social Survey (GSS)](http://www.gss.norc.org/) gathers data on American society in order to monitor and explain trends in attitudes, behaviours, and attributes. Many trends have been tracked for decades, so one can see the evolution of attitudes, etc in American Society.


In this blog post, we analyze data from the **2016 GSS sample data**, using it to estimate values of *population parameters* of interest about US adults. The GSS sample data file has 2867 observations of 935 variables, but we are only interested in very few of these variables - therefore we will use a sample that contains only time spend in `email`, `snapchat`, `instagram`, `twitter`; their sex; and what level of education they have.

Let's take a look at the data

```{r, read_gss_data, cache=TRUE, echo=TRUE, include=FALSE}
gss <- read_csv(here::here("datasets", "smallgss2016.csv"), 
                na = c("", "Don't know",
                       "No answer", "Not applicable"))

head(gss, 10)
```

You will also notice that many responses should not be taken into consideration, like "No Answer", "Don't Know", "Not applicable", "Refused to Answer". These were classified as "`NA`"

## Instagram and Snapchat, by sex

Can we estimate the *population* proportion of Snapchat or Instagram users in 2016? Let's find out.

### Data loading, cleaning and wrangling

We will first proceed to clean this dataset a bit and do data wrangling. We will create a new variable that is *Yes* if the respondent reported using any of Snapchat (`snapchat`) or Instagram (`instagrm`), and *No* if not. We will leave the `NA` and drop them later on.

```{r snap_insta, echo=TRUE}

snap_insta <- gss %>%
  
  # Create new variable
  mutate(snap_insta = case_when (
    instagrm=="Yes" | snapchat=="Yes" ~ "Yes",
    instagrm=="No" & snapchat=="No" ~ "No",
    # Add NA as string
    TRUE ~ "NA" )) 

# Show results
head(snap_insta,5)
```

### Calculating proportions
We will calculate the proportion of Yes’s for `snap_insta` among those who answered the question, i.e. excluding NAs.

```{r yes_prop, echo=TRUE}

yes_prop <-snap_insta %>% 
  # Filter out NA
  filter(!snap_insta=="NA") %>% 
  # Create proportion of yes
  summarise(yes = sum(snap_insta=="Yes"), # Count those who said "Yes"
            count = n()) %>%  # Count all records
  # Calculate proportion (x/n)
  mutate(yes_prop = yes/count)

# Show results
kable(yes_prop)
```
Our point estimate (proportion) of people with snapchat or instagram seems to be about 37.5%.

### Confidence intervals

We will first use the CI formula for proportions to construct 95% CIs for men and women who used either Snapchat or Instagram. To do this, we will need to group records by sex and calculate the corresponding summary statistics (proportion and standard error)

```{r ci_formula_snap_insta, echo=TRUE}
# Create new object "df" (short for data frame)
df <-snap_insta %>% 
  # Filter out NA
  filter(!snap_insta=="NA") %>%
  # Group by gender
  group_by(sex) %>%
  # Calculate summary statistics
  summarise(yes = sum(snap_insta=="Yes"),
            count = n()) %>%
  # Calculate proportion and standard error of the proportion sample
  mutate(yes_prop = yes/count,
        se = sqrt(yes_prop*(1-yes_prop)/count))

# Show results
kable(df)

```

We see that the proportion of women who use snapchat or instagram seems to be of about 42%, whereas that of men is 39%. We wil now establish a confidence interval of 95% using the conventional formula. Since the distribution of the proportion is binomial, we can approach the sample with a normal distribution and calculate the proportion estimate to infer the population parameter of the proportion.

```{r confidence_interval_df, echo=TRUE}
# We have a proportion, yet we create a CI around the proportion. The proportion is of the sample.

df_ci<-df %>%
  group_by(sex) %>%
  # Establish lower limit
  summarise(lower= yes_prop - (1.96*se),
            # Establish upper limit
            upper= yes_prop + (1.96*se),
            yes_prop=yes_prop,
            se=se) %>%
  select(sex, lower, yes_prop, upper, se)

kable(df_ci)
```

As it seems, the women's confidence interval goes from 38% to 44%, with the estimate proportion around 42%. Men's interval goes from 28% to 35%, with the estimate proportion around 32%. We can appreciate this on the following plot.

```{r ci_df_plot}

# We will use ggerrorplot from the library ggpubr
ggerrorplot(df_ci, 
            x ="sex", 
            y = "yes_prop", 
            color = "sex",
            desc_stat = "mean_ci", 
            palette="uchicago", 
            size=.9) +
  
  # Add bar of CI
  geom_errorbar(aes(ymin = lower, 
                    ymax = upper), width = .2, position=position_dodge(.9) ) +
  
  # Add labels
  labs(title = "Do more women use instagram/snapchat than men?",
       subtitle = "Figure 2.1",
       y= "Estimate proportion of people who use instagram/snapchat",
       x= "Gender",
       caption = "Source: General Social Survey (GSS)") +
  
  # Remove legend
  theme(legend.position = "none")


```

If our results could be generizable, then we could say that in proportion, more women use instragram/snapchat than men. There is a statistical significance in the difference between the proportions, as the confidence intervals do not overlap.

## Twitter, by education level

Can we estimate the *population* proportion of Twitter users by education level in 2016?. 

### Data loading, cleaning and wrangling

There are 5 education levels in variable `degree` which, in ascending order of years of education, are Lt high school, High School, Junior college, Bachelor, Graduate. 

We will first turn `degree` from a character variable into a factor variable. We will convert this into a factor and establish the levels so the plot will take those in a certain order.

```{r twitter_df, echo=TRUE}

# Convert degree into a factor and establish levels
twitter_df <- gss %>%
  mutate(degree = factor(degree, c("Lt high school","High School","Junior college","Bachelor","Graduate")))

# Show results

kable(head(twitter_df, 5))
```

Now we will create a  new variable, `bachelor_graduate` that is *Yes* if the respondent has either a `Bachelor` or `Graduate` degree. As before, if the recorded value for either was NA, the value will also be NA. We will drop these for the analysis as well.

```{r bachelor_graduate, echo=TRUE}
twitter_df <- twitter_df %>% 
  # Create bachelor_graduate
  mutate(bachelor_graduate = case_when(
    degree %in% c("Bachelor","Graduate") ~ "Yes",
    is.na(degree) ~ "NA",
    
    #Else
    TRUE ~ "No"))

# Filter out NAs
twitter_df <- twitter_df %>% filter(!bachelor_graduate=="NA")

# Show Results
kable(head(twitter_df, 10))

```
Our data set is now good to go.

### Calculating proportions

We will calculate the proportion of `bachelor_graduate` who do (Yes) and who don't (No) use twitter. 


```{r prop_bach_grad, echo=TRUE}
# Create new object
prop_twitter_df <- twitter_df %>%
   # filter by Yes in bachelor graduates
  filter(bachelor_graduate=="Yes" & !twitter =="NA") %>%
  # Group by Yes/No
  group_by(twitter) %>%
  # Count records
  summarise(count=n()) %>%
  # Create total count and proportion for each group
  mutate(total = sum(count),
         prop = count/total) %>%
  # Show "Yes" first
  arrange(desc(twitter))

# Show results
kable(head(prop_twitter_df, 10))
```



Using the CI formula for proportions, we will construct two 95% CIs for `bachelor_graduate` vs whether they use (Yes) and don't (No) use twitter. 


```{r ci_formula}
# Create confidence interval with formula

prop_twitter_df<- prop_twitter_df %>%
  # Create standard error, lower and upper
  mutate(se = sqrt(prop*(1-prop)/total),
         lower = prop - se*1.96,
         upper = prop + se*1.96) %>%
  # Select relevant variables
  select(twitter, total, count, se, lower, prop, upper)

# Show results
kable(head(prop_twitter_df, 10))
```

According to our estimates, 76% of those who are educated up to either bachelor/masters degree do use twitter, whereas 24% are not. The confidence interval for those who are goes from 73% to 80%, and that of those who are not goes from 19% to 27%. As we can appreciate in the following plot, the intervals **DO NOT** overlap.

```{r plot_intervals_twitter, echo=TRUE}

# Use ggpubr for this 
ggerrorplot(prop_twitter_df, 
            x ="twitter", 
            y = "prop", 
            color = "twitter",
            desc_stat = "mean_ci", 
            palette="uchicago", size=.9)+
  # Establish interval reach
  geom_errorbar(aes(ymin = lower, ymax = upper),width = .2, position=position_dodge(.9))+
  
  # Add labels
  labs(title = "University gradutes do not twitter around",
       subtitle = "Figure 2.2",
       y= "Propotions of bachelor/graduates who use or not twitter",
       x= "Bachelor/Graduates",
       caption = "Source: General Social Survey (GSS)") +
  theme(legend.position = "none")


```
**Is twitter not appealing for university educated people?**

Confidence intervals clearly do not overlap. There is a significant difference between one another. Therefore, there is a statistical significance between the bachelor/graduates and the rest of the people when it comes to using twitter. Does this mean that bachelors/graduates do not find twitter appealing? One hypothesis would be that twitter is more suitable for teenagers. Data on the activity of users and age would be useful to analyse this.

## Email usage

Can we estimate the *population* parameter on time spent on email weekly?

### Data loading, cleaning and wrangling

We will go back to our initial dataset and create a new variable called `email` that combines `emailhr` and `emailmin` to reports the number of minutes the respondents spend on email weekly.

```{r df_email, echo=TRUE}
# Create new object

df_email <- gss %>%
  # Add minutes for those that are not NA
  mutate(email = case_when(
    # Use as.numeric since those variables are "char" format.
    emailhr!="NA" & emailmin !="NA" ~ (as.numeric(emailhr)*60 + as.numeric(emailmin)))) %>%# Multiply hour by 60 to convert into minutes
    # Filter out "NA"
    filter(!email=="NA")
# Show results

kable(head(df_email, 10))
```

### Distribution

Let's visualise the distribution of this new variable and find the mean and the median number of minutes respondents spend on email weekly.

```{r email_dist, echo=TRUE}
# Create plot distribution

ggplot(df_email, aes(x=email)) +
  # Use histogram
  geom_histogram(color = 'white', bins=40) +
  # Add labels
  labs(title="Americans and Emails - how much time do they spend? ",
       subtitle="Figure 2.3 | Distribution of minutes spend on email by American Students",
       y="Count",
       x="Minutes on emails",
       caption= "Source: General Social Survey (GSS)")

# Show mean and median
df_mean_median <- df_email %>%
  summarise(mean = mean(email),
            median = median(email))

kable(df_mean_median)

```

Based on the plots, we can appreciate that the distribution of weekly minutes spend on email is `right-skewed`. Because of this, the mean will not be an interesting measure technique, as it will be completely sensitive to the outliers on the right. Therefore we should proceed to use the median to analyse it.


### Bootstrap and CI
Using the `infer` package, we will calculate a 95% bootstrap confidence interval for the mean amount of time Americans spend on email weekly. 

```{r infer_email, echo=TRUE}
# Set seed

set.seed(1234)
# Simulate bootstrap
boot_email <- df_email %>%
  # Specify variable
  specify(response = email) %>%
  # Simulation
  generate(reps = 1000, type = "bootstrap") %>%
  # Calculate median
  calculate(stat='mean')

# Distribution of bootstrap median
ggplot(boot_email, aes(x=stat)) +
  geom_histogram(color='white') +
  labs(title = "Mean sample bootstrap distribution",
     subtitle = "Figure 2.4 | The mean sample is normally distributed as expected",
     caption = "Source: General Social Survey (GSS)",
     y= "Sample mean variation")

# Confidence interval
ci_boot <-boot_email %>%
  get_confidence_interval(level=0.975, type='percentile')

ci_boot

# Mean
mean_time <-boot_email %>%
  summarise(mean = mean(stat))

# Convert mean into readable hours and minutes 
time <- hms::hms(sec=0, min=mean_time$mean, hours=0)

# Convert lower and upper into readable hours and minutes
time_lower<- hms::hms(sec=0, min=ci_boot$lower_ci, hours=0)
time_upper <-hms::hms(sec=0, min=ci_boot$upper_ci, hours=0)

# Paste results
print(paste0("Confidence interval of 95% includes a lower of: ",hour(time_lower)," hours and ",minute(time_lower), " minutes." ), quote=FALSE)
print(paste0("Time spent on average weekly on email: ",hour(time)," hours and ",minute(time), " minutes." ), quote=FALSE)
print(paste0("Confidence interval of 95% includes an upper of: ",hour(time_upper)," hours and ",minute(time_upper), " minutes." ), quote=FALSE)
```

According to our results, on average Americans might spend around almost 7 hours per week on the email, with a confidence interval lower of 6 hours and 20 minutes and upper of 7 hours and almost 40 minutes. 

If we analysed this with a 99% confidence interval, the results would come as follows:

```{r}
boot_email %>%
  get_confidence_interval(level=0.99, type='percentile')
```

A 99% confidence interval will be wider than a 95% confidence interval because to be more confident that the true population value falls within the
interval we will need to allow more potential values within the interval. In that case, the z-score would equal 2.58 (as opposed to 1.96)

```{r 99_CI}

# Z score with 95% confidence
qnorm(0.975)
# Z score with 99% confidence
qnorm(0.995)
```

In most times, the cost is not worth the less of variability.

#### Was this the right approach?

The problem with the last bit is that the distribution was heavily `right-skewed` - therefore, as mentioned before, the mean might not be a good estimate of the population parameter as it will be sensitive to outliers. However, if we wanted to estimate the sample median on bootstrapping, we would get the following: 

```{r sample_median, echo=TRUE}


# Set seed
set.seed(1234)
# Simulate bootstrap
boot_email <- df_email %>%
  # Specify variable
  specify(response = email) %>%
  # Simulation
  generate(reps = 1000, type = "bootstrap") %>%
  # Calculate median
  calculate(stat="median")

# Distribution of bootstrap median
ggplot(boot_email, aes(x=stat)) +
  geom_histogram(color='white') +
  labs(title = "Sample median bootstrap distribution",
     subtitle = "Figure 2.5 | Uniform distribution for the median",
     caption = "Source: General Social Survey (GSS)",
     y= "Sample median variation")

# Confidence interval
ci_boot <-boot_email %>%
  get_confidence_interval(level=0.975, type='percentile')

ci_boot

# Mean
mean_time <-boot_email %>%
  summarise(mean = median(stat))

# Convert mean into readable hours and minutes 
time <- hms::hms(sec=0, min=mean_time$mean, hours=0)

# Convert lower and upper into readable hours and minutes
time_lower<- hms::hms(sec=0, min=ci_boot$lower_ci, hours=0)
time_upper <-hms::hms(sec=0, min=ci_boot$upper_ci, hours=0)

# Paste results
print(paste0("Confidence interval of 95% includes a lower of: ",hour(time_lower)," hours and ",minute(time_lower), " minutes." ), quote=FALSE)
print(paste0("Time spent on average weekly on email: ",hour(time)," hours and ",minute(time), " minutes." ), quote=FALSE)
print(paste0("Confidence interval of 95% includes an upper of: ",hour(time_upper)," hours and ",minute(time_upper), " minutes." ), quote=FALSE)


```
#### Why is this happening?

The median would have a `uniform` distribution. The problem here is that too many fall in or near zero.

```{r zero_values, echo=TRUE}
# How many people are NOT using the email?
df_email %>% filter(email==0) %>% summarise(count=n())

```
217 People do NOT use the email at all during the week in our sample. That is about 13% of the sample size.

If we check how many use it less than 100:

```{r less_100, echo=TRUE}
# How many people barely use the email
df_email %>% filter(email<100) %>% summarise(count=n())

```
We are talking about 720 people who barely use the email.

#### Solving this issue

In this analysis we will just exclude those who do not use it at all `(!email==0)` - excluding a bigger sample with the idea that "they barely use it" would be infering too much in our data, so we will just stick to excluding those with 0 minutes. Let's analyse the whole thing again.


```{r exclude_no_users, echo=TRUE}

# Set seed
set.seed(1234)
# Simulate bootstrap
boot_email <- df_email %>%
  # Filter no-users
  filter(!email==0) %>%
  # Specify variable
  specify(response = email) %>%
  # Simulation
  generate(reps = 1000, type = "bootstrap") %>%
  # Calculate median
  calculate(stat="median")

# Distribution of bootstrap median
ggplot(boot_email, aes(x=stat)) +
  geom_histogram(color='white') +
  labs(title = "Sample median bootstrap distribution",
     subtitle = "Figure 2.6 | Uniform distribution for the median",
     caption = "Source: General Social Survey (GSS)",
     y= "Sample median variation")

# Confidence interval
ci_boot <-boot_email %>%
  get_confidence_interval(level=0.975, type='percentile')

ci_boot

# Mean
median_time <-boot_email %>%
  summarise(median = median(stat))

# Convert mean into readable hours and minutes 
time <- hms::hms(sec=0, min=median_time$median, hours=0)

# Convert lower and upper into readable hours and minutes
time_lower<- hms::hms(sec=0, min=ci_boot$lower_ci, hours=0)
time_upper <-hms::hms(sec=0, min=ci_boot$upper_ci, hours=0)

# Paste results
print(paste0("Confidence interval of 95% includes a lower of: ",hour(time_lower)," hours and ",minute(time_lower), " minutes." ), quote=FALSE)
print(paste0("Time spent (median) weekly on email: ",hour(time)," hours and ",minute(time), " minutes." ), quote=FALSE)
print(paste0("Confidence interval of 95% includes an upper of: ",hour(time_upper)," hours and ",minute(time_upper), " minutes." ), quote=FALSE)


```
As a result, we get a not so much uniform distribution, with a minimum of 2 hours, a median of 2 hours and a maximum of 3 hours.